---
output: 
  pdf_document: 
    keep_tex: yes
---
**A General Boosting Regression Model (GMB) to Predict a Weightlifting Exercise**   
  
**By: Wahsabii Neanderthal**  

**Purpose.** To predict the manner in which subjects did a barbell curl exercise.  
  
**Data.**  
The data used were from six subjects who participated in a dumbbell lifting exercise. The subjects used the dumbbell five different ways. The five ways, as described in the study, were "exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E). Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. The classe variable is the response variable to predict. 

```{r Data, warning=FALSE, message=FALSE}
# load library used to model data
library(caret)

# Read data file into memory
filename <- "pml-training.csv"  # Assumes user has preloaded data in working directory
pmlData <- read.csv(filename, header=TRUE, na.strings=c("NA","#DIV/0!",""))
pmlData <- pmlData[,complete.cases(t(pmlData))] # eliminate incomplete columns 
pmlData <- pmlData[-c(1:7)] # remove the first seven admin columns 

# Show the structure of the data
str(pmlData)
```
To create the model, the data was partitioned in two segments with approximately 60% of the data used to tune the model and the remaining 40% used to validate the model.
```{r BuildModel, warning=FALSE, message=FALSE}
# Partition the data 
set.seed(8675309)
inTrain <- createDataPartition(y=pmlData$classe, p=0.60, list=FALSE)
training <- pmlData[inTrain,]
testing <-  pmlData[-inTrain,]
c(dim(training), dim(testing))
```
A density plot of the classe data factors are as follows:  
```{r DensityPlot, warning=FALSE, message=FALSE}
# Density plot of the classe factor variable
qplot(classe, colour=classe, data=training, geom="density")
```

**How the model was built.**   
The Generalized Boosting Regression (gmb) Model Package in r's caret package was used to model the various sensor readings recorded in pmlData above.  The classe variable is the response variable and all the other sensor data were explanatory variables.   
  
**How cross validation was used.**  
Repeated cross validation was used to model pmlData. Three folds and three repeats were used to model the training data.   

**Expected out of sample error estimate.**  
Initial hypothesis for the expected out of error rate was 5 to 10% was acceptable.   

**Why model choices were made.** 
Three folds and three repeats instead of 10 were chosen due to the excessively long computing time to build a model. 

```{r LoadModel, warning=FALSE, message=FALSE}
# WARNING: this took me 12 hours to build this model
# if the gbmFit2 model exist load else build it
if(file.exists("gbmModel.Rda")){load("gbmModel.Rda")} else {
          
        fitControl <- trainControl(## 3-fold CV
                method = "repeatedcv",
                number = 3, 
                repeats = 3)
        
        gbmGrid <-  expand.grid(interaction.depth = c(1, 5, 9),
                                n.trees = (1:30)*50,
                                shrinkage = 0.1,
                                n.minobsinnode = 20)
        set.seed(8675309)
        gbmFit2 <- train(classe ~ ., data = training,
                         method = "gbm",
                         trControl = fitControl,
                         verbose = FALSE,
                         tuneGrid = gbmGrid)
}
```
The model exceeded expectations at 99% accuracy. All twenty test questions were predicted with 100% accuracy. 
```{r Prediction, warning=FALSE, message=FALSE}
# Out of sample prediction
testPred <- predict(gbmFit2,testing)
# Accuracy
confusionMatrix(testPred,testing$classe)
```
